Approximation as Differentiation

### 2025-05-24 main/formal-structures.tex
\GPS macro to standardise terminology.

We take an approach based on the CHAD (Combinatory Homomorphic Automatic Differentiation) of Mattias Vákár and others~\cite{vákár22,nunes2023}. Here follows a rough overview of this approach and how we have applied it to Galois program slicing. \todo{Final version shouldn't be ``rough''!}

@@

For Galois program slicing, we again take $\cat{C}$ to be $\Set$ and now take $T(X) = X \to \LatGal$: indexed collections of lattices, whose elements are interpreted as approximations at the given point, with indexed Galois connections as the maps between them.

@@

First we introduce the category of lattices and Galois connections, which will serve as our base category for
interpreting forward and backward Galois slicing. Galois connections are pairs of monotone functions $f: Y \to
X$ and $g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse
of $g$, and $g$ the best approximation from above to an inverse of $f$. This setup is a nice fit for the
bidirectional nature of (dynamic) program slicing, with $f$ capturing \emph{backward} slicing (producing the
least input slice for a given slice of the output), and $g$ capturing \emph{forward} slicing (producing the
greatest output slice for a given slice of the input).

@@

As sketched in \secref{introduction:galois-slicing}, for Galois slicing we want a setting where the
approximants of a point $x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$
representing the approximation that discards all information about $x$, greatest element $\top$ representing
the approximation that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways
to combine approximations. Thus rather than working directly in $\PosGal$, we consider the following
subcategory instead.

>>>>

We take an approach based on the CHAD (Combinatory Homomorphic Automatic Differentiation) of Mattias Vákár and others~\cite{vákár22,nunes2023}. Here follows a rough overview of this approach and how we have applied it to \GPS.

@@

For \GPS, we again take $\cat{C}$ to be $\Set$ and now take $T(X) = X \to \LatGal$: indexed collections of lattices, whose elements are interpreted as approximations at the given point, with indexed Galois connections as the maps between them.

@@

First we introduce the category of lattices and Galois connections, which will serve as our base category for
interpreting forward and backward \GPS. Galois connections are pairs of monotone functions $f: Y \to
X$ and $g: X \to Y$ between posets, where $f$ is the (pointwise) best approximation from below to an inverse
of $g$, and $g$ the best approximation from above to an inverse of $f$. This setup is a nice fit for the
bidirectional nature of (dynamic) program slicing, with $f$ capturing \emph{backward} slicing (producing the
least input slice for a given slice of the output), and $g$ capturing \emph{forward} slicing (producing the
greatest output slice for a given slice of the input).

@@

As sketched in \secref{introduction:galois-slicing}, for \GPS we would like a setting where the approximants
of a point $x$ form a bounded lattice $(X, \meet, \join, \top, \bot)$, with least element $\bot$ representing
the approximation that discards all information about $x$, greatest element $\top$ representing the
approximation that retains all information about $x$, and $\meet$ and $\join$ providing two canonical ways to
combine approximations. Thus rather than working directly in $\PosGal$, we consider the following subcategory
instead.

### 2025-05-24 main/framework.tex
Correction.
\GPS macro to standardise terminology.

\section{A categorical framework for Galois slicing}

@@

Like any functor category $\Func{X}{\cat{C}}$ inherits limits and colimits from its codomain.

@@

\subsection{Galois slicing via the Grothendieck construction}

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for Galois slicing. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every
$x \in X$, a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all
pairs $(f, \partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial
f(x): \partial X(x) \to \partial Y(f(x))$, as we sketched informally at the beginning of this section. The
pullback (reindexing) along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f,
\reindex{\partial g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois
connection at each point, using the baseline (unapproximated) function $f$.

>>>>

\section{A categorical framework for \GPS}

@@

Like any functor category $\Func{X}{\cat{C}}$ inherits limits and colimits pointwise from its codomain.

@@

\subsection{\GPS via the Grothendieck construction}

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for \GPS. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every $x \in X$,
a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all pairs $(f,
\partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial f(x): \partial
X(x) \to \partial Y(f(x))$, as we sketched informally at the beginning of this section. The pullback
(reindexing) along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f,
\reindex{\partial g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois
connection at each point, using the baseline (unapproximated) function $f$.

### 2025-05-24 main/introduction.tex
\GPS macro to standardise terminology.

  Now suppose that for a given run of the program, we are interested in which of the numerical parts of the input are used to compute the output for the query parameters $l = \mathsf{a}$ and $l = \mathsf{b}$. We can use Galois Program Slicing to do this. We arrange for the approximations of the input to form the following lattice, where the actual piece of data is at the top and information lost by approximation is represented by $\bot$s:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      % row 2
      \node [below of=top] (ioi) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [left of=ioi,xshift=-3cm] (oii) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      \node [right of=ioi,xshift=3cm] (iio) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      % row 3
      \node [below of=ioi] (oio) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      \node [left of=oio,xshift=-3cm] (ooi) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [right of=oio,xshift=3cm] (ioo) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};
      % row 4
      \node [below of=oio] (bot) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};

@@

  Galois program slicing associates with each run of the program a Galois connection telling us how the inputs and outputs are related in that run. The backwards portion $\partial (\mathit{query}\,l)_r$ tells us, given an approximation of the output, what the least approximation of the input is needed to generate that output. In the case of the two runs considered in this example, if we say we are not interested in the output by feeding in the least approximation $\bot$, then we find that we only need the least approximation of the input:
  \begin{displaymath}
    \partial (\mathit{query}\,l)_r(\bot) = [(\mathsf{a},\bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]
  \end{displaymath}
  for both $l = \mathsf{a}$ and $l = \mathsf{b}$. If instead we take the greatest approximation of the output (i.e., the output ``$1$'' itself), then the two query runs' backwards approximations return different results:
  \begin{displaymath}
    \begin{array}{l}
      \partial (\mathit{query}\,\mathsf{a})_r(1) = [(\mathsf{a},0), (\mathsf{b},\bot), (\mathsf{a},1)] \\
      \partial (\mathit{query}\,\mathsf{b})_r(1) = [(\mathsf{a},\bot), (\mathsf{b},1), (\mathsf{a},\bot)]
    \end{array}
  \end{displaymath}
  Pieces of the input that were {\em not} used are replaced by $\bot$. As we expect, the run of the query with label $\mathsf{a}$ depends on the entries in the database labelled with $\mathsf{a}$, and likewise for the run with label $\mathsf{b}$.

@@

  In a simple query like this, it is easy to work out the dependency relationship between the input and output. However, the benefit of Galois Program Slicing is that it is {\em automatic} for all programs, no matter how complex the relationship between input and output. Moreover, by changing what we mean by ``approximation'' we can compute a range of different information about a program.
\end{example}

Previous work on Galois program slicing uses an operational semantics of programs that generates a trace of each execution which can be interpreted after the fact to compute the Galois connection described above. This becomes {\em program slicing} by including the source code of the program as part of the input, so that, in the backward direction, the least approximation of the input required for an output approximation includes the least part of the program required.

@@

\subsection{Galois Program Slicing and Automatic Differentiation}

\todo{explain why we want a more abstract view on Galois program slicing}
There is a close analogy between Galois Program Slicing and Automatic Differentiation for differentiable programs. We have already hinted at this in the description above, but let us now make it explicit.

\begin{itemize}
\item For Galois Program Slicing, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For Galois Program Slicing, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For Galois Program Slicing, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For Galois Program Slicing, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}

Given this close connection between Galois program slicing and differentiable programming, we can take structures intended for modelling automatic differentiation and use them to model Galois program slicing. This will enable us to generalise and expand the scope of Galois program slicing to act as a foundation for data provenance in a wide range of computational settings.

@@

Our primary contribution in this article is to elucidate Galois Program Slicing by relating it to differentiable programming and automatic differentiation. In doing so, we aim to expand the applicability of program slicing, and to potentially transfer efficient automatic differentiation implementation techniques from automatic differentiation to program slicing and provenance tracking.

\begin{enumerate}
\item We explain how the CHAD framework of Vákár et al. can be adapted to give a general categorical framework for Galois Program Slicing,
\item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of Galois Program Slicing.
\item With the benefit of our abstract setting, we can relate Galois Program Slicing to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.
\item We also relate Galois Program Slicing to categorical treatments of differentiability via Tangent Categories \note{Bob: not sure about this.}
\end{enumerate}

>>>>

  Now suppose that for a given run of the program, we are interested in which of the numerical parts of the input are used to compute the output for the query parameters $l = \mathsf{a}$ and $l = \mathsf{b}$. We can use \GPS to do this. We arrange for the approximations of the input to form the following lattice, where the actual piece of data is at the top and information lost by approximation is represented by $\bot$s:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      % row 2
      \node [below of=top] (ioi) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [left of=ioi,xshift=-3cm] (oii) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      \node [right of=ioi,xshift=3cm] (iio) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      % row 3
      \node [below of=ioi] (oio) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      \node [left of=oio,xshift=-3cm] (ooi) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [right of=oio,xshift=3cm] (ioo) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};
      % row 4
      \node [below of=oio] (bot) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};

@@

  \GPS associates with each run of the program a Galois connection telling us how the inputs and outputs are related in that run. The backwards portion $\partial (\mathit{query}\,l)_r$ tells us, given an approximation of the output, what the least approximation of the input is needed to generate that output. In the case of the two runs considered in this example, if we say we are not interested in the output by feeding in the least approximation $\bot$, then we find that we only need the least approximation of the input:
  \begin{displaymath}
    \partial (\mathit{query}\,l)_r(\bot) = [(\mathsf{a},\bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]
  \end{displaymath}
  for both $l = \mathsf{a}$ and $l = \mathsf{b}$. If instead we take the greatest approximation of the output (i.e., the output ``$1$'' itself), then the two query runs' backwards approximations return different results:
  \begin{displaymath}
    \begin{array}{l}
      \partial (\mathit{query}\,\mathsf{a})_r(1) = [(\mathsf{a},0), (\mathsf{b},\bot), (\mathsf{a},1)] \\
      \partial (\mathit{query}\,\mathsf{b})_r(1) = [(\mathsf{a},\bot), (\mathsf{b},1), (\mathsf{a},\bot)]
    \end{array}
  \end{displaymath}
  Pieces of the input that were {\em not} used are replaced by $\bot$. As we expect, the run of the query with label $\mathsf{a}$ depends on the entries in the database labelled with $\mathsf{a}$, and likewise for the run with label $\mathsf{b}$.

@@

  In a simple query like this, it is easy to work out the dependency relationship between the input and output. However, the benefit of \GPS is that it is {\em automatic} for all programs, no matter how complex the relationship between input and output. Moreover, by changing what we mean by ``approximation'' we can compute a range of different information about a program.
\end{example}

Previous work on \GPS uses an operational semantics of programs that generates a trace of each execution which can be interpreted after the fact to compute the Galois connection described above. This becomes {\em program slicing} by including the source code of the program as part of the input, so that, in the backward direction, the least approximation of the input required for an output approximation includes the least part of the program required.

@@

\subsection{Galois Slicing and Automatic Differentiation}

\todo{explain why we want a more abstract view on \GPS}
There is a close analogy between \GPS and automatic differentiation for differentiable programs. We have already hinted at this in the description above, but let us now make it explicit.

\begin{itemize}
\item For \GPS, we assume that every value has an associated lattice of {\em approximations}. For differentiable programs, every point has an associated vector space of {\em tangents}.
\item For \GPS, every program has an associated forward approximation map that takes approximations forward from the input to the output. This map {\em preserves meets}. For differentiable programs, every program has a forward derivative that takes tangents of the input to tangents of the output. The forward derivative map is {\em linear}, so it preserves addition of tangents and the zero tangent.
\item For \GPS, every program has an associated backward approximation map that takes approximations of the output back to least approximations of the input. This map {\em preserves joins}. For differentiable programs, every program has a reverse derivative that takes tangents of the output to tangents of the input. This map is again {\em linear}.
\item For \GPS, the forward and backward approximation maps are related by being a Galois connection. For differentiable programming, the forward and reverse derivatives are related by being each others' transpose.
\end{itemize}

Given this close connection between \GPS and differentiable programming, we can take structures intended for modelling automatic differentiation and use them to model \GPS. This will enable us to generalise and expand the scope of \GPS to act as a foundation for data provenance in a wide range of computational settings.

@@

Our primary contribution in this article is to elucidate \GPS by relating it to differentiable programming and automatic differentiation. In doing so, we aim to expand the applicability of program slicing, and to potentially transfer efficient automatic differentiation implementation techniques from automatic differentiation to program slicing and provenance tracking.

\begin{enumerate}
\item We explain how the CHAD framework of Vákár et al. can be adapted to give a general categorical framework for \GPS,
\item Using our adapted CHAD framework, we can explain how type structure can be used to control the approximation lattices associated with data points, something that was ``hard coded'' in previous presentations of \GPS.
\item With the benefit of our abstract setting, we can relate \GPS to other parts of denotational semantics. In particular, we show that there is a close connection with Stable Domain Theory, a proposed solution to capturing sequentiality by recording more intensional information about programs' sensitivity to approximations.
\item We also relate \GPS to categorical treatments of differentiability via Tangent Categories \note{Bob: not sure about this.}
\end{enumerate}

### 2025-05-24 notes/stability.tex
\GPS macro to standardise terminology.

Stable functions are an antecedent of Galois slicing that arose in
domain theory as an attempt to capture the idea that any function
describing a computational process must explore its input in a way
that allows identification of minimal parts of the input that are
needed for chosen approximations of the output.

>>>>

Stable functions are an antecedent of \GPS that arose in
domain theory as an attempt to capture the idea that any function
describing a computational process must explore its input in a way
that allows identification of minimal parts of the input that are
needed for chosen approximations of the output.
